Bioconductor workflow for single-cell RNA sequencing: Dimensionality reduction, clustering, and lineage inference

### 2017-06-27 workflow/workflow.Rmd
Edits to text.

# Analysis of olfactory stem cell lineages using scRNA-seq data

@@

## Dimensionality reduction adjusting for batches: ZINB-WaVE

@@

[SD: Include Fig. 1 from [@Risso2017]?]
As with most dimensionality reduction methods, the user needs to specify the number of dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for the batch effects.

@@

When function `zinbDimRed` is called, normalized values of the counts can be computed and used for visualization (e.g. in heatmaps). The normalized values are the residuals of the fit of the ZINB-WaVE model where the gene and cell covariates were specified by the user when calling function `zinbDimRed`, but no low dimensional matrix is computed. In that case, the residuals represent the counts adjusted for the gene and cell covariates and should capture the biological signal of interest. To compute the residuals, the deviance is used.

@@

As expected, the normalized values are uniformly distributed for the different batches. See Figure \@ref(fig:boxplotNorm).

@@

The principal component analysis (PCA) of the normalized values shows that, as expected, cells do not cluster by batch but by orginal published clusters. See Figure \@ref(fig:pcanorm). Overall, looking at the normalized values, it seems that the batch effect has been removed.

@@

Calling function `zinbDimRed` from package `zinbwave`, we performed dimensionality reduction where the number of dimensions in the low dimensional space is `K = 50`. We call the matrix in the low dimensional space `W`. We can visualize `W` in two dimensions by performing multidimensional scaling (MDS) on the distance matrix of `W` where the euclidian distance is used. To make sure that `W` captured the biological signal of interest, we plotted the MDS of `W` with colors corresponding to the original published clusters. See Figure \@ref(fig:mdsW).

>>>>

# Analysis of olfactory stem cell differentiation using scRNA-seq data

@@

## Dimensionality reduction: ZINB-WaVE

@@

[SD: Include Fig. 1 from [@Risso2017]?]
As with most dimensionality reduction methods, the user needs to specify the number of dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for batch effects.

@@

The function `zinbDimRed` returns normalized expression measures as deviance residuals from the fit of the ZINB-WaVE model with user-supplied gene- and cell-level covariates. Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots). Note that in this case no low-dimensional matrix is computed, as some of the biological signal of interest could be removed when computing residuals with respect to this matrix.

@@

As expected, the normalized values no longer exhibit batch effects (Figure \@ref(fig:boxplotNorm)).

@@

The principal component analysis (PCA) of the normalized values shows that, as expected, cells do not cluster by batch but by the orginal clusters (Figure \@ref(fig:pcanorm)). Overall, it seems that normalization was effective at removing batch effects.

@@

The `zinbDimRed` function can also be used to perform dimensionality reduction, where the user-supplied dimension `K` of the low-dimensional space is set here to `K = 50`. The resulting low-dimensional matrix `W` can be visualized in two dimensions by performing multi-dimensional scaling (MDS) using the Euclidian distance. To verify that `W` indeed captured the biological signal of interest, we plotted the MDS of `W` with colors corresponding to the original published clusters (Figure \@ref(fig:mdsW)).
